Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2023-12-19 paper.tex
Fix formatting

\section{Introduction}
Structures, or spatial arrangements of matter and energy, including some fields (e.g., velocity or pressure) are ubiquitous in research applications and frequently require description for subsequent analysis. Examples of such objects for studies include galaxy formations \cite{springel2006}, immiscible multiphase fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples \cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010}, food specimens \cite{derossi2019,nagdalian2021}, biological tissue \cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and interstellar turbulence \cite{portillo2018developing}. The scale of interest may span orders of magnitude from angstroms and nm for materials \cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can be non-static -- it will  change with time, and this dynamics is usually also of great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous bounds and such;
  \item Prediction of material physical properties using machine (deep) learning \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison and analysis \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular}, which may include temporal dynamics or surface evolution \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023} in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature extraction), and retrieval (by means of stochastic reconstructions) \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off between field-of-view and imaging resolution \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

@@

The large portion of the research papers mentioned above \cite{rozenbaum2014,karsanina2015,ledesma2018,derossi2019,portillo2018developing,takada2003three,hopkins2013stars,jiao2013,Torquato_book,roding2020predicting,KarsaninaEJSS,capek2009,gerke2019tensor,REVpaper,LavrukhinPRE,Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,jiao2007,SciRep1,Havelka,Geoderma2018} is based on so-called correlation functions (CFs). These functions define a probability that some events are occurring withing the structure (with some special cases of surface functions), for example, 2-point probability $S_2$ measures the probability that both ends of the line segment are within the same phase, 2-point cluster function $C_2$ - that both ends of the line segment are within the same interconnected cluster of same phase, and lineal path $L_2$ measures the probability that the whole segment lies within the phase (see \highlight{Fig.1}). Compared to other descriptors, CFs possess a number of features making them, in our opinion, a perfect candidate for universal descriptor:
\begin{enumerate}
  \item They contain the majority of other metrics inside. For example, porosity is $S_2(0)$ (and at zero correlation length of the majority of correlation functions except for surface ones), surface area is represented by in the derivative at $S_2(0)$ \cite{debye1957scattering} or the value of $F_{ss}(0)$ (or $F_{sss}(0)$). Connectivity is described by cluster functions $C_2$ or $C_3$. Pore-size distribution is related to pore-size function and partially by $L_2$ and $S_3$. Surface-void functions are interconnected with mean curvature and Euler characteristic \cite{ma2020generation}. All together, the classic set of correlation functions describes both the geometry and topology of the structure at hand;
  \item While the extent to which such classical set describes any arbitrary structure is not yet known, compared to other complex descriptors  where are known pathways to assess the information content of correlation functions -- there is a methodology for $S_2$ \cite{Gommes2} and approaches for other functions \cite{Degeneraty.045306,cherkasov2023towards};
  \item In addition to a possibility to compute correlation functions using images or other structural data, i.e., similar to all other descriptors, some CFs can be also determined experimentally \cite{debye1957scattering,barrall1992nmr,dietrich1995scattering,li2018accurate};
  \item Correlation functions can be scaled \cite{karsaninaPRL} or manipulated, for example, to add functions into a single one \cite{moctezuma2002};
  \item As opposed to scalar metrics such as porosity, surface area or Minkowski functionals, CFs can be computed in directions and describe anisotropic structures;
  \item Numerous rigorous bounds to obtain thermal, elastic, electromagnetic and transport properties are known, among which the most known (and least useful due to the impractically huge gap between the bounds) are Hashin-Strikman bounds \cite{hashin1963variational}. Other expressions based on integration of higher order $S_n$ are known and were found to be close to direct simulations \cite{brown1955solid,beran1965use,milton1981bounds,hlushkou2015effective};
  \item One can control both the size of the CFs-based descriptor and the information content via correlation length, number of functions in the set (including directions \cite{EPL2}), and by the order of functions (e.g., 2- and 3-point versions, or even more points).
\end{enumerate}
This list highlights that correlations functions are indeed possess necessary properties so satisfy key criteria stated above.

@@

We mentioned that classic set of correlation functions describes
both the geometry and topology of the structure at hand - but this is to yet unknown extent.
With the advent of analytical methods \cite{cherkasov2023towards} to assess information content of arbitrary descriptor for structure of any size, it will be possible to choose the most useful set of correlation functions and compare their information content against classical and other metrics.

>>>>

\section{Introduction}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

@@

The large portion of the research papers mentioned above
\cite{rozenbaum2014,karsanina2015,ledesma2018,derossi2019,portillo2018developing,
  takada2003three,hopkins2013stars,jiao2013,Torquato_book,roding2020predicting,
  KarsaninaEJSS,capek2009,gerke2019tensor,REVpaper,LavrukhinPRE,Adler_recon,Y-T,
  tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,jiao2007,SciRep1,Havelka,Geoderma2018}
is based on so-called correlation functions (CFs). These functions define a
probability that some events are occurring withing the structure (with some
special cases of surface functions), for example, 2-point probability $S_2$
measures the probability that both ends of the line segment are within the same
phase, 2-point cluster function $C_2$ -- that both ends of the line segment are
within the same interconnected cluster of same phase, and lineal path $L_2$
measures the probability that the whole segment lies within the phase (see
\highlight{Fig.1}). Compared to other descriptors, CFs possess a number of
features making them, in our opinion, a perfect candidate for universal
descriptor:
\begin{enumerate}
  \item They contain the majority of other metrics inside. For example, porosity
    is $S_2(0)$ (and at zero correlation length of the majority of correlation
    functions except for surface ones), surface area is represented by in the
    derivative at $S_2(0)$ \cite{debye1957scattering} or the value of
    $F_{ss}(0)$ (or $F_{sss}(0)$). Connectivity is described by cluster
    functions $C_2$ or $C_3$. Pore-size distribution is related to pore-size
    function and partially by $L_2$ and $S_3$. Surface-void functions are
    interconnected with mean curvature and Euler characteristic
    \cite{ma2020generation}. All together, the classic set of correlation
    functions describes both the geometry and topology of the structure at hand;
  \item While the extent to which such classical set describes any arbitrary
    structure is not yet known, compared to other complex descriptors where are
    known pathways to assess the information content of correlation functions --
    there is a methodology for $S_2$ \cite{Gommes2} and approaches for other
    functions \cite{Degeneraty.045306,cherkasov2023towards};
  \item In addition to a possibility to compute correlation functions using
    images or other structural data, i.e., similar to all other descriptors,
    some CFs can be also determined experimentally
    \cite{debye1957scattering,barrall1992nmr,dietrich1995scattering,li2018accurate};
  \item Correlation functions can be scaled \cite{karsaninaPRL} or manipulated,
    for example, to add functions into a single one \cite{moctezuma2002};
  \item As opposed to scalar metrics such as porosity, surface area or Minkowski
    functionals, CFs can be computed in directions and describe anisotropic
    structures;
  \item Numerous rigorous bounds to obtain thermal, elastic, electromagnetic and
    transport properties are known, among which the most known (and least useful
    due to the impractically huge gap between the bounds) are Hashin-Strikman
    bounds \cite{hashin1963variational}. Other expressions based on integration
    of higher order $S_n$ are known and were found to be close to direct
    simulations
    \cite{brown1955solid,beran1965use,milton1981bounds,hlushkou2015effective};
  \item One can control both the size of the CFs-based descriptor and the
    information content via correlation length, number of functions in the set
    (including directions \cite{EPL2}), and by the order of functions (e.g., 2-
    and 3-point versions, or even more points).
\end{enumerate}
This list highlights that correlations functions are indeed possess necessary
properties so satisfy key criteria stated above.

@@

We mentioned that classic set of correlation functions describes
both the geometry and topology of the structure at hand -- but this is to yet
unknown extent. With the advent of analytical methods
\cite{cherkasov2023towards} to assess information content of arbitrary
descriptor for structure of any size, it will be possible to choose the most
useful set of correlation functions and compare their information content
against classical and other metrics.
